Demes: a specification for interchanging complex demography

### 2022-04-05 paper.tex
Expanded on selfing, clonal rates

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little or no programming effort.

@@

\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
	and why that is required for the standard coalescent.}

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilisation or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilisation given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection.
Thus, the history of population size changes does not depend on mean fitnesses within
a population nor on differences in mean fitness between populations.
This indepedence of population size and
selection is often called ``soft" selection \citep{christiansen1975hard}.

@@

\begin{figure}[h!]
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The compact non-redundant
        representation expressed using YAML. (B) The same model in the fully-qualified
        form. (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
    }
\end{figure}

>>>>

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\jkcomment}[1]{{\textcolor{red}{JK: #1}}}
\newcommand{\ggcomment}[1]{{\textcolor{yellow!60!red}{GG: #1}}}
\newcommand{\krtcomment}[1]{{\textcolor{purple}{KRT: #1}}}
\newcommand{\mhcomment}[1]{{\textcolor{cyan}{MH: #1}}}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information. \mhcomment{Maybe add a short, concrete example to convince readers of the problem?}
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

@@

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab:software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little programming effort.

@@

\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
	and why that is required for the standard coalescent.}
\mhcomment{The `exchangeability' assumption also breaks down under different reproductive modes, since the sampling of the alleles affects coalescent times. For example, in a highly selfing species the coalescent time for two different alleles within an individual is much lower than two alleles from different individuals. I've tried to rewrite the below paragraph to update and clarify the reproduction parameters in line with the discussions on Github. Finally, 'rates' imply events occur as Poisson-like processes over continuous time, whereas the values used here instead define the \emph{fraction} of reproductions of each type. Shall we rename to selfing/clonal `fractions', or would that mess up the rest of the schema too much?}

Each population has a selfing rate and cloning rate assigned to it, where each denotes the probability that offspring are generated from one generation to the next by either self-fertilisation or cloning of an individual. More specifically, denote the clonal rate by $\sigma$ and the selfing rate by $S$. In each population a proportion of offspring $\sigma$ reproduce clonally, while $1-\sigma$ reproduce sexually. Within the sexually--reproducing offspring, $S$ were born via self--fertilisation while the rest were outcrossed. Hence $S$ and $\sigma$ can take any value between zero and one, but do not have to sum to one. See \citet{hartfield2016facsexcoal} for more details of the mathematical properties of these rates in a coalescent context. \mhcomment{If requested I can quickly draw up a schematic diagram to illustrate these partitions.} Note that even if $S$, $\sigma$ are set to zero then some residual selfing can occur due to sampling in the Wright--Fisher model.

%Within a population, individuals are assumed to be exchangeable, but there are
%also parameters for nonrandom mating, such as selfing or cloning rates, which
%give the probability that offspring are generated from one generation to the
%next by self-fertilisation or cloning of an individual. Cloning and selfing
%rates can take any value between zero and one, and if both are specified then
%the selfing rate is the probability an offspring is generated through
%self-fertilisation given it was produced sexually (i.e. with probability one
%minus the cloning rate).

We omit any details regarding selection.
Thus, the history of population size changes does not depend on mean fitnesses within
a population nor on differences in mean fitness between populations.
This indepedence of population size and
selection is often called ``soft" selection \citep{christiansen1975hard}.

@@

%\begin{figure}[h!]
%    \begin{minipage}{0.44\textwidth}
%        \begin{tcolorbox}
%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
%        \end{tcolorbox}
%        \begin{tcolorbox}
%            \includegraphics[width=\textwidth]{fig/IM}
%        \end{tcolorbox}
%    \end{minipage}\hfill
%    \begin{minipage}{0.54\textwidth}
%        \begin{tcolorbox}
%            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
%        \end{tcolorbox}
%    \end{minipage}\\
%    \caption{
%        \label{fig:IM}
%        Example isolation-with-migration Demes model. (A) The compact non-redundant
%        representation expressed using YAML. (B) The same model in the fully-qualified
%        form. (C) A visual representation of the model using \texttt{demesdraw}.
%        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
%        but are much more human-readable.
%    }
%\end{figure}
